---
title: "Interactive Cancer Incidence Visualizations"
output: html_document
---

## Dataset Description

This dashboard uses data from the publicly available SEER colorectal cancer dataset, provided by the National Cancer Institute.  
Data were collected from hospital records and state cancer registries covering individuals aged 18 to 50 residing in Georgia.  
The dataset includes case counts and population data aggregated by county and race group.  
The analysis focuses on the time period from 2000 to 2010.  
County-level incidence rates were calculated using standardized population counts and matched to geographic shapefiles for visualization.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Widget 1: Interactive Scatter Plot

This scatter plot shows the relationship between **population size** and **cancer incidence rate** in Georgia counties. You can hover over points to see detailed data, and use zoom/pan features for exploration.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(plotly)

# ==== Step 1: Load and aggregate incidence data ====

data_file <- "E:/Project/Data/processed_filtered_male_2000.txt"
if (!file.exists(data_file)) stop("Data file not found.")

incidence_raw <- read.table(data_file, header = TRUE, sep = "\t",
                            stringsAsFactors = FALSE, fileEncoding = "UTF-8")

# Clean population column
incidence_raw$Population <- as.numeric(gsub(",", "", incidence_raw$Population))

# Remove invalid population rows
incidence_raw <- incidence_raw[!is.na(incidence_raw$Population) & incidence_raw$Population > 0, ]

# Group by county, aggregate count and population
incidence_data <- incidence_raw %>%
  group_by(GEOID, county_name) %>%
  summarise(
    Total_Pop = sum(Population),
    Total_Count = sum(Count),
    .groups = "drop"
  ) %>%
  mutate(
    Incidence_Rate = (Total_Count / Total_Pop) * 100000
  )

# Ensure numeric field for x-axis
incidence_data$Total_Pop <- as.numeric(incidence_data$Total_Pop)

# ==== Step 2: Create interactive plotly scatter plot ====

p1 <- plot_ly(
  incidence_data, 
  x = ~Total_Pop, 
  y = ~Incidence_Rate, 
  text = ~paste0(
    "County: ", county_name, "<br>",
    "Population: ", formatC(Total_Pop, format = "d", big.mark = ","), "<br>",
    "Incidence Rate: ", sprintf("%.2f", Incidence_Rate), " per 100K"
  ),
  type = 'scatter', 
  mode = 'markers',
  marker = list(size = 10, color = 'blue'),
  hoverinfo = "text",
  width = NULL,
  height = NULL
) %>%
  layout(
    autosize = TRUE,
    title = list(
      text = "Incidence Rate vs Population",
      y = 0.95
    ),
    margin = list(t = 100),
    xaxis = list(title = "Total Population"),
    yaxis = list(title = "Incidence Rate per 100,000")
  ) %>%
  config(responsive = TRUE)

p1

```

### Key Takeaways:
- The majority of counties have populations under 500,000 and incidence rates between 5 and 20 per 100,000.
- Larger counties show less variability in incidence rates, clustering around consistent values.
- Counties with smaller populations tend to have greater variation, including outliers with very high rates, possibly due to data instability in low-population areas.

---

## Widget 2: Interactive Incidence Map

This map visualizes **colorectal cancer incidence rates** across **Georgia counties**. The color intensity represents the rate per 100,000 people. You can hover over each county for more details.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(sf)
library(leaflet)
library(dplyr)
library(htmltools)

# ==== Step 1: Load and aggregate incidence data ====

data_file <- "E:/Project/Data/processed_filtered_male_2000.txt"
if (!file.exists(data_file)) stop("Data file not found.")

incidence_raw <- read.table(data_file, header = TRUE, sep = "\t",
                            stringsAsFactors = FALSE, fileEncoding = "UTF-8")

# Clean population column
incidence_raw$Population <- as.numeric(gsub(",", "", incidence_raw$Population))

# Remove invalid population rows
incidence_raw <- incidence_raw[!is.na(incidence_raw$Population) & incidence_raw$Population > 0, ]

# Group by county, aggregate count and population
incidence_data <- incidence_raw %>%
  group_by(GEOID, county_name) %>%
  summarise(
    Total_Pop = sum(Population),
    Total_Count = sum(Count),
    .groups = "drop"
  ) %>%
  mutate(
    Incidence_Rate = (Total_Count / Total_Pop) * 100000
  )

# Ensure GEOID is character
incidence_data$GEOID <- as.character(incidence_data$GEOID)

# ==== Step 2: Load and merge shapefile ====

shp_file <- "E:/Project/Data/tl_2010_13_county10.shp"
if (!file.exists(shp_file)) stop("Shapefile not found.")

georgia_counties <- st_read(shp_file)
georgia_counties$GEOID10 <- as.character(georgia_counties$GEOID10)

# Merge shapefile and processed incidence data
g_map_data <- left_join(georgia_counties, incidence_data, by = c("GEOID10" = "GEOID"))

# Remove counties without incidence data
g_map_data <- g_map_data[!is.na(g_map_data$Incidence_Rate), ]

# ==== Step 3: Define hover tooltip and color scale ====

labels <- sprintf(
  "<strong>County:</strong> %s<br/>
   <strong>Total Population:</strong> %s<br/>
   <strong>Incidence Rate:</strong> %.2f per 100K",
  g_map_data$county_name,
  formatC(g_map_data$Total_Pop, format = "d", big.mark = ","),
  g_map_data$Incidence_Rate
) %>% lapply(htmltools::HTML)

pal <- colorNumeric("Blues", domain = g_map_data$Incidence_Rate, na.color = "transparent")

# ==== Step 4: Draw interactive leaflet map ====

leaflet(g_map_data, options = leafletOptions(zoomControlPosition = "bottomleft")) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~pal(Incidence_Rate),
    weight = 1,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 3,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "13px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~Incidence_Rate,
    opacity = 0.7,
    title = "Incidence Rate<br>(per 100,000)",
    position = "bottomright"
  ) %>%
  addControl(
    html = "<div style='font-size:10px;'><strong>Colorectal Cancer Incidence in Georgia</strong></div>",
    position = "topright"
  )
```

### Key Takeaways:
- Colorectal cancer incidence is not evenly distributed across Georgia’s counties.
- Several counties in central and southwestern Georgia display noticeably higher incidence rates.
- The spatial distribution suggests that localized factors may influence colorectal cancer risk and deserve further investigation.

---

## Source Code
The full source code is embedded in this document and available for review. The visualizations are generated using **plotly** and **tmap**, ensuring interactivity and usability.

<p style="font-size:16px">
🔗 View project source code on <a href="https://github.com/rocinante0v0/Interactive-Cancer-Incidence-Visualizations" target="_blank">GitHub</a>
</p>

## Real-World Impact

This dashboard helps identify Georgia counties with unusually high colorectal cancer incidence rates, guiding targeted public health action.  
It provides a visual tool to support resource allocation, community outreach, and the investigation of geographic health disparities.


**End of Report.**
